Higher-Order unification for free!

### 2024-04-22 paper/main.tex
gzip: page breaking seems now good

For the detection of \notllambda it
is sufficient to identify applications with a flexible head and arguments that
are not in distinct names. The following rule for \notllambda compilation
is inserted  just before rule~\ref{rule:compapp}.

@@

Given a \linkbeta $l$ of the form \linkbetaM{\Gamma}{T}{\rhsBeta}, we provide
the following activation rules:

@@

We will relax this last failure condition to accommodate for the
implementation of heuristics in~\cref{sec:heuristics}.

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R HolesInR, !,
  declare_constraint (link-eta L R) [L|HolesInR].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
\end{elpicode}
% \todo{improve code}
% link-eta L R :- not (maybe-eta R), !, contract R S, L = S.

@@

Different strategies can be used to unify terms of the object language.
% in the meta langauge. 
The first approach that comes to mind consist in implementing \Uo as a
regular routine in the ML, i.e. write rules as follows:

\begin{elpicode}
  decision X :- X ~\Uo~(all A x\ app [P, x]), finite A,
    pi x\ decision (app [P, x]).
\end{elpicode}

\noindent
This method fails to take advantage of the logic
programming engine provided by the ML, as it degrades indexing by
eliminating all data from rule heads. Additionally,
implementing a unification procedure in the ML is likely to be significantly
slower compared to the built-in one.

Another possibility is to avoid having application and abstraction nodes
in the syntax tree, and use the ones of the ML, as in:

@@

\noindent
However, this encoding has two big limitations. First it is
not always feasible to adopt it for Coq due to the fact that the type system
of the ML is too limited to accommodate for that one the object language,
e.g. Coq can typecheck variadic functions~\cite{cpdt}. \\
Second, the encoding of Coq terms provided by Elpi is primarily utilized for meta
programming, in order to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor} and \texttt{arg}. In this context, constants need to live in
an open world, akin to the \elpiIn{string} data type used in the preceding
examples, they cannot be symbols of the meta language.

@@

Another work that is, somewhat surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.\\
~\\
\emph{The approach presented in this paper}
%provides a third option that 
addresses all the concerns
mentioned earlier. It takes advantage of the
unification capabilities of the meta language
at the price of handling ``problematic'' sub terms on
the side.\\
As a result %of this choice 
our encoding takes
advantage of indexing data structures and
mode analysis for clause filtering.
It it worth mentioning that we only replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable by the meta language logic programming engine.
Some preliminary benchmarks on the Stdpp
library~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018} show
encouraging results: our type-class solver has some
overhead on small goal but is consistently faster than
the Coq one starting from goals made of XXX nodes.\\
\todo{D: how many?}
Our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment.
Moreover, the unification process we obtain is ready to take
advantage of potential improvements to the programming engine,
such as tabled search. Finally, our shallow encoding can
enable to apply forms of static analysis for
the meta language, such as determinacy, also to the object language.

>>>>

By inserting this rule
before rule~\ref{rule:compapp}.
we identify terms \notllambda, i.e. flexible applications with arguments that
are not distinct names:

@@

%We add the followins rules to \elpiIn{progress1}.
Let $l$ be a \linkbeta of the form \linkbetaM{\Gamma}{T}{\rhsBeta}

@@

We relax this condition and accommodate for
% the implementation of 
heuristics in~\ref{sec:heuristics}.

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R HolesInR, !,
  declare_constraint (link-eta L R) [L|HolesInR].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
\end{elpicode}
\todo{check progress-rhs is back}

@@

Object-language terms can be unified using
different strategies.

The first approach that comes to mind consist in implementing \Uo as a
regular routine, i.e. write rules as follows

\begin{elpicode}
  decision X :- unify X (all A x\ app [P, x]), finite A,
    pi x\ decision (app [P, x]).
\end{elpicode}

\noindent
where \elpiIn{unify} is a regular predicate.
This method fails to take advantage of the logic
programming engine provided by the meta language since
it removes all data from rule's heads and makes
indexing degrade. Additionally,
implementing a unification procedure in the meta language is likely to be significantly
slower compared to the built-in one.

Another possibility is to avoid having application and abstraction nodes
in the syntax tree, and use the meta language ones, as in:

@@

\noindent
However, this encoding has two big limitations. First it is
not always feasible to adopt it for Coq due to the fact that the type system
of the meta language is too limited to accommodate for that one the object language,
e.g. Coq can typecheck variadic functions~\cite{cpdt}. \\
Second, the encoding of Coq terms provided by Elpi is primarily used for meta
programming, i.e. to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor/3} and \texttt{arg/3}. To avoid that constants 
cannot be symbols of the meta language, but must rather
live in
an open world, akin to the \elpiIn{string} data type used in the \elpiIn{con}
constructor.

@@

Another work that is, somewhat surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.\\
~\\
\emph{The approach presented in this paper}
%provides a third option that 
addresses all the concerns
mentioned earlier. It takes advantage of the
unification capabilities of the meta language
at the price of handling problematic sub terms on
the side.
As a result %of this choice 
our encoding takes
advantage of indexing data structures and
mode analysis for clause filtering.
It is worth mentioning that we only replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable. % by the meta language logic programming engine.
Some preliminary benchmarks on the Stdpp
library~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018} show
encouraging results: our type-class solver has some
overhead on small goals but is consistently faster than
the Coq one starting from goals made of XXX nodes.\\
\todo{D: how many?}
Our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment.
Moreover, the unification process we obtain is ready to take
advantage of potential improvements to the programming engine,
such as tabled search. Finally, our shallow encoding can
enable to apply forms of static analysis for
the meta language, such as determinacy, also to the object language.
